A formal treatment of bidirectional typing

### 2023-10-27 tex/ESOP24/BiSig.tex
[ tex ] Update Demonstration
Update BiSig.tex

@@

\title{A formal treatment of bidirectional typing}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for al signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

@@

%\begin{acks}
%%We thank Nathanael Arkor for the useful conversation.
%We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
%The work is supported by the \grantsponsor{NSTC}{National Science and Technology Council of Taiwan}{https://www.nstc.gov.tw/} under grant \grantnum{NSTC}{NSTC 112-2221-E-001-003-MY3}.
%\end{acks}

@@

%\input{appendix}
\subfile{S7-formalisation}

>>>>

\numberwithin{table}{section}

@@

\title{A formal treatment of bidirectional typing%
%  \thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

@@

@@

\subfile{S7-formalisation}
\input{appendix}
